Communicative reduction in referring expressions within a multi-player negotation game

### 2023-01-30 write-ups/cogsci23_paper/manuscript.Rmd
fix typos

abstract: >
    The ability to form novel conventions is a key signature of efficient linguistic conmmunication. Reduction of referring expressions, one measure of convention formation, is found robustly in dyadic repeated reference games when the target images are initially difficult to name. In reference games, participants share the explicit goal of establishing joint reference. However, establishing reference is a key subgoal of many conversations where interlocuters have more complex goals. In the current work, we analyze a dataset where reference was embedded in strategic 3- player negotiation and coordination games. In these more complex games, we found that the patterns of reduction and convergence to conventions still held across two different incentive conditions, with some suggestive differences between the conditions.

@@

Successful communication is grounded in a shared understanding of what utterances mean in the context of their production. Whether the communicative goals are cooperative or adversarial, interlocuters need to establish joint reference to be effective. In many cases, conventional word meanings are enough, but there are also contexts where objects to be referred to are not easy to distinguish. In these situations, establishing reference is more difficult, but no less important.

The formation of ad-hoc referring expressions has been studied extensively in the context of dyadic reference games [@clarkReferringCollaborativeProcess1986; @krauss1964changes]. In these, two participants see a set of images (often abstract shapes) and one person is tasked with identifying an image so the other person can pick it out. People can be very successful at this task, achieving high levels of matches. Over repeated reference to the same images, pairs develop shorthand names for the images, leading to shorter descriptions in later repetitions. These nicknames conventionalize with a pair, as they tend to stick to the same description, but are idiosyncratic and differ between pairs. This task has been a well-studied microcosm for understanding reference which has proven useful for testing theories about how referring expressions originate, how expressions are designed, and when they change [@brennan1996conceptual; @metzing2003conceptual; @leung2020parents; @weber2003; @yoonAudienceDesignMultiparty2019].

@@

We build on work by @mankewitz2021 who created a task with the aim of bridging the gap between reference games with hard-to-name objects and negotiation datsets with easily nameable objects. They created a 3-person game where players each selected what flowers to grow from a set of 6 flowers with variable and partially hidden value. Players got the value of the flower only if they were the only one to select it, which incentivized coordination and negotiation. However, all the flowers shown were the same color, so they were not easily nameable, and the flowers repeated, leading to opportunities for ad-hoc conventionalization. This game structure encouraged players to describe flowers while never explicitly telling a player to refer to any specific flower. The experiment had two conditions, one where players within a group won points together (and thus had fully aligned incentives) and another where they won points as individuals (and didn't). @mankewitz2021 found a slight decrease in language over the course of the games, but they did not conduct a detailed analysis to assess whether the decreases in words were associated with shorter and more conventionalized referring expressions. Here, we address this gap by separating out the referring expressions in order to check for patterns of convention-formation.

Our work addresses whether participants in a complex, multi-player strategic game with differing incentives and shared knowledge are able to develop conventionalized reference expressions. We explore whether the phenomena found in classic dyadic iterated reference games extends to this more naturalistic domain, and whether the results differ across the individual and shared incentive conditions. The key questions we addressed were:

@@

Historically, studies of convention formation have had to rely on proxies such as reduction, subjective judgements of similarity, and and manual measures of lexical overlap. In contrast, we use natural language processing tools for mapping sentences to semantic vector spaces to quantify semantic similarities between sentences, following recent work by @hawkins2020 and @boyce2022.

@@

## Procedure
As described in @mankewitz2021, participants played a real-time coordination game in groups of three, implemented using Empirica [@almaatouqEmpiricaVirtualLab2020]. On each trial, each group saw a set of 6 flower images (Figure \ref{interface}A) . Each participant saw the values for 4 of the flowers (represented as a colored bar), such that each flower's value was hidden from one participant. Players could coordinate and discuss using a chat box before each selected a flower. If one player selected a flower, it was worth the shown reward; if multiple players collided and selected the same flower, they each got a lower reward instead (Figure \ref{interface}B). Players needed to communicate about the flowers in order to coordinate their selections. The rewards translated in a monetary bonus for the participants at the end of the game.

@@

## Participants
@mankewitz2021 recruited `r 3 *(d.games |> filter(conditionName!="Old Cooperative") |> nrow())` participants for 25 3-person games in each of the individual and shared utilities conditions. Some games did not complete the full 24 trials because participants dropped out early.  We excluded incomplete games from analyses, which left us with `r d.games.final |> filter(condition=="Individual Utilities") |> nrow()` games in the individual utilities condition and `r d.games.final |> filter(condition=="Shared Utilities") |> nrow()` in the shared utilities condition (`r 3*(d.games.final |> nrow())` participants included in the analyses).

@@

Two games consisting of `r raw_rows` utterances were annotated by both annotators.  `r nrow(both)` utterances were identified as containing reference expressions by at least one annotator; of these annotators agreed on the exact reference expression for `r agree_span` (`r round(agree_span/nrow(both)*100)`%) of the cases. The second annotator only coded the target of the referring expression for `r nrow(count)` of the utterances. Of these, the two annotators agreed on the target in `r agree_target` (`r round(agree_target/nrow(count)*100)`%) of cases. We take this level of interannotator agreement as an indication that the reference spans and targets were identified reliably. We extracted a total of `r annotated |> filter(numword>0) |>  tally()` referring expressions.

## SBERT embeddings
Following recent work by @hawkins2020 and @boyce2022, we used tools for natural langauge processing to quantify similarity between phrases in order to assess the dynamics of convergence to conventions.

We embedded each of the extracted referring expressions using SBERT [@reimers2019] which maps each utterance to a vector in a high-dimensional semantic space. We use cosine distance between pairs of vector embeddings as a measure of similarity between the corresponding referring expressions.  Higher cosine values correspond to vectors close together, and thus, reference expressions with more similar meanings. Some examples of pairs of referring expressions and their similarities are shown in Table \ref{costable}.

@@

The theory of partner-specificity predicts that descriptions to the same flower in different groups will diverge over time as each group forms their own distinct convention. Table \ref{extable} shows how groups descriptions of the same flower may diverge (flower 2), but also how groups might choose related but different descriptions (flower 3).

@@

### Within games
Games in the shared utilities condition had non-significantly more similar descriptions of the same flower within games (`r stats_text(wi_g_wi_f,2)`), but this did not interact with trial number (`r stats_text(wi_g_wi_f,3)`) or whether the utterances were produced by the same participant (`r stats_text(wi_g_wi_f,5)`). Games in the shared utilities condition did have lower similarities between descriptions of different flowers within games (`r stats_text(wi_g_bet_f,2)`), but this was moderated by trial (`r stats_text(wi_g_bet_f,3)`) and whether the utterances were produced by the same player (`r stats_text(wi_g_bet_f,5)`).

Comparing to the end conventions, games in the shared utility condition produced flower descriptions that were more similar to the convention for that flower (`r stats_text(to_e_wi_f,2)`), but condition did not interact with trial number (`r stats_text(to_e_wi_f,3)`) and did not significantly interact with same participant versus different participant (`r stats_text(to_e_wi_f,5)`).  In the shared utility condition, descriptions were slightly more similar to the conventions for other flowers (`r stats_text(to_e_bet_f,2)`), but condition did not interact with trial (`r stats_text(to_e_bet_f,3)`). This similarity was even higher if the description and convention came from the same person (`r stats_text(to_e_bet_f,5)`). As a whole, groups in the shared utility may have had more internal alignment on the mapping between flowers and names, but this pattern was inconsistent.

@@

Lastly, we can look at how the conventionalized names compared, both between members of the same group and across players in different groups that saw the same color palette of flowers, and whether these comparisons differ by condition (Figure \ref{withinend}).

We treated different flowers described by different games as the baseline condition. For individual utility games, descriptions from groupmates were more similar (even for different flowers, `r stats_text(wi_e,8)`), descriptions of the same flower were more similar (even across games, `r stats_text(wi_e,6)`), and there was a large interaction effect, where descriptions of the same flower were very similar among groupmates (`r stats_text(wi_e,7)`).

@@

These findings are suggestive that in the shared utility condition, games differentiated less from each other, and their conventions may have been influenced more by shared priors for how to describe the flowers than the conventions in individual utility games.

@@

First, we saw reduction, as referring expressions tended to decrease in length over the course of the game. Second, we found the expected set of convergence and divergence patterns. References to the same flower within a game converged both to other descriptions used at that timepoint and to the post-game descriptions. This convergence was specific to flower and games, as references to different flowers within a game and references to the same flower across games both diverged.

@@

Our third consideration was how the different utility conditions moderated the reduction and convergence effects. We found some slight and hard to interpret condition differences within games. In addition, groups in the shared utility condition diverged less from each other. This pattern of condition differences is consistent with less idiosyncratic descriptions in the shared utility condition.

@@

This dataset is just the start of testing how well findings from reference games generalize to more complex conversational situations. One dimension that we were unable to explore here due to the design of the game is the idea of shared versus privileged knowledge [@keysar2000taking; @heller2008role]. In the flowers game, like in reference games, all players saw the same set of images (and knew they did), which makes making descriptions to flowers easier. In real-world situations, interlocuters may have different sets of objects in their context and may not know what is in their interlocuter's context. An important future direction is studying how repeated reference phenomena interact with different communcation goals and contexts.

@@

\setlength{\parindent}{-0.1in} 
\setlength{\leftskip}{0.125in}
\noindent
\small

>>>>

abstract: >
    The ability to form novel conventions is a key signature of efficient linguistic communication. Reduction of referring expressions, one measure of convention formation, is found robustly in dyadic repeated reference games when the target images are initially difficult to name. In reference games, participants share the explicit goal of establishing joint reference. However, establishing reference is a key subgoal of many conversations where interlocuters have more complex goals. In the current work, we analyze a dataset where reference was embedded in strategic 3- player negotiation and coordination games. In these more complex games, we found that the patterns of reduction and convergence to conventions still held across two different incentive conditions, with some modest differences between the conditions.

@@

Successful communication is grounded in a shared understanding of what utterances mean in the context of their production. Whether the communicative goals are cooperative or adversarial, interlocutors need to establish joint reference to be effective. In many cases, conventional word meanings are enough, but there are also contexts where objects to be referred to are not easy to distinguish. In these situations, establishing reference is more difficult, but no less important.

The formation of ad-hoc referring expressions has been studied extensively in the context of dyadic reference games [@clarkReferringCollaborativeProcess1986; @krauss1964changes]. In these, two participants see a set of images (often abstract shapes) and one person is tasked with identifying an image so the other person can pick it out. People can be very successful at this task, achieving high levels of matches. Over repeated reference to the same images, pairs develop shorthand names for the images, leading to shorter descriptions in later repetitions. These nicknames conventionalize within each pair, as they tend to stick to the same description, but are idiosyncratic and differ between pairs. This task has been a well-studied microcosm for understanding reference which has proven useful for testing theories about how referring expressions originate, how expressions are designed, and when they change [@brennan1996conceptual; @metzing2003conceptual; @leung2020parents; @weber2003; @yoonAudienceDesignMultiparty2019].

@@

We build on work by @mankewitz2021 who created a task with the aim of bridging the gap between reference games with hard-to-name objects and negotiation datasets with easily nameable objects. They created a 3-person game where players each selected what flowers to grow from a set of 6 flowers with variable and partially hidden value. Players got the value of the flower only if they were the only one to select it, which incentivized coordination and negotiation. However, all the flowers shown were the same color, so they were not easily nameable, and the flowers repeated, leading to opportunities for ad-hoc conventionalization. This game structure encouraged players to describe flowers while never explicitly telling a player to refer to any specific flower. The experiment had two conditions, one where players within a group won points together (and thus had fully aligned incentives) and another where they won points as individuals (and didn't). @mankewitz2021 found a slight decrease in language over the course of the games, but they did not conduct a detailed analysis to assess whether the decreases in words were associated with shorter and more conventionalized referring expressions. Here, we address this gap by separating out the referring expressions in order to check for patterns of convention-formation.

Our work addresses whether participants in a complex, multi-player strategic game with differing incentives and shared knowledge are able to develop conventionalized reference expressions. We explore whether the phenomena found in classic dyadic iterated reference games extend to this more naturalistic domain, and whether the results differ across the individual and shared incentive conditions. The key questions we addressed were:

@@

Historically, studies of convention formation have had to rely on proxies such as reduction, subjective judgments of similarity, and manual measures of lexical overlap. In contrast, we use natural language processing tools for mapping sentences to semantic vector spaces to quantify semantic similarities between sentences, following recent work by @hawkins2020 and @boyce2022.

@@

## Procedure
As described in @mankewitz2021, participants played a real-time coordination game in groups of three, implemented using Empirica [@almaatouqEmpiricaVirtualLab2020]. On each trial, each group saw a set of 6 flower images (Figure \ref{interface}A) . Each participant saw the values for 4 of the flowers (represented as a colored bar), such that each flower's value was hidden from one participant. Players could coordinate and discuss using a chat box before each player selected a flower. If one player selected a flower, it was worth the shown reward; if multiple players collided and selected the same flower, they each got a lower reward instead (Figure \ref{interface}B). Players needed to communicate about the flowers in order to coordinate their selections. The rewards translated into a monetary bonus for the participants at the end of the game.

@@

## Participants
@mankewitz2021 recruited `r 3 *(d.games |> filter(conditionName!="Old Cooperative") |> nrow())` participants total for 25 3-person games in each of the individual and shared utilities conditions. Some games did not complete the full 24 trials because participants dropped out early.  We excluded incomplete games from analyses, which left us with `r d.games.final |> filter(condition=="Individual Utilities") |> nrow()` games in the individual utilities condition and `r d.games.final |> filter(condition=="Shared Utilities") |> nrow()` in the shared utilities condition (`r 3*(d.games.final |> nrow())` participants included in the analyses).

@@

Two games consisting of `r raw_rows` utterances were annotated by both annotators.  `r nrow(both)` utterances were identified as containing reference expressions by at least one annotator; of these, the annotators agreed on the exact reference expression for `r agree_span` (`r round(agree_span/nrow(both)*100)`%) of the cases. The second annotator only coded the target of the referring expression for `r nrow(count)` of the utterances. Of these, the two annotators agreed on the target in `r agree_target` (`r round(agree_target/nrow(count)*100)`%) of cases. We take this level of inter-annotator agreement as an indication that the reference spans and targets were identified reliably. We extracted a total of `r annotated |> filter(numword>0) |>  tally()` referring expressions.

## SBERT embeddings
Following recent work by @hawkins2020 and @boyce2022, we used tools for natural language processing to quantify similarity between phrases in order to assess the dynamics of convergence to conventions.

We embedded each of the extracted referring expressions using SBERT [@reimers2019] which maps each utterance to a vector in a high-dimensional semantic space. We used cosine distance between pairs of vector embeddings as a measure of similarity between the corresponding referring expressions.  Higher cosine values correspond to vectors close together, and thus, reference expressions with more similar meanings. Some examples of pairs of referring expressions and their similarities are shown in Table \ref{costable}.

@@

The theory of partner-specificity predicts that descriptions to the same flower in different groups will diverge over time as each group forms their own distinct convention. Table \ref{extable} shows how descriptions of the same flower may diverge (flower 2), but also how groups might choose related but different descriptions (flower 3).

@@

### Within games
Games in the shared utilities condition had non-significantly more similar descriptions of the same flower within games (`r stats_text(wi_g_wi_f,2)`), but this did not interact with trial number (`r stats_text(wi_g_wi_f,3)`) or whether the utterances were produced by the same participant (`r stats_text(wi_g_wi_f,5)`). Games in the shared utilities condition did have lower similarities between descriptions of different flowers within games (`r stats_text(wi_g_bet_f,2)`), but this was moderated by trial (`r stats_text(wi_g_bet_f,3)`) and by whether the utterances were produced by the same player (`r stats_text(wi_g_bet_f,5)`).

Comparing to the end conventions, games in the shared utility condition produced flower descriptions that were more similar to the convention for that flower (`r stats_text(to_e_wi_f,2)`), but condition did not interact with trial number (`r stats_text(to_e_wi_f,3)`) and did not significantly interact with same participant versus different participant (`r stats_text(to_e_wi_f,5)`).  In the shared utility condition, descriptions were slightly more similar to the conventions for other flowers (`r stats_text(to_e_bet_f,2)`), but condition did not interact with trial (`r stats_text(to_e_bet_f,3)`). This similarity was even higher if the description and convention came from the same person (`r stats_text(to_e_bet_f,5)`). As a whole, groups in the shared utility condition may have had more internal alignment on the mapping between flowers and names, but this pattern was inconsistent.

@@

### Comparison of end conventions
Lastly, we looked at how the conventionalized names compared, both between members of the same group and across players in different groups that saw the same color palette of flowers, and whether these comparisons differed by condition (Figure \ref{withinend}).

We treated different flowers described by different games as the baseline condition. For individual games, descriptions from groupmates were more similar (even for different flowers, `r stats_text(wi_e,8)`), descriptions of the same flower were more similar (even across games, `r stats_text(wi_e,6)`), and there was a large interaction effect, where descriptions of the same flower were very similar among groupmates (`r stats_text(wi_e,7)`).

@@

These findings are suggestive that in the shared utility condition, games differentiated less from each other, and their conventions may have been influence more by shared priors for how to describe the flowers more than the conventions in individual utility games were.

@@

First, we saw reduction, as referring expressions tended to decrease in length over the course of the game. Second, we found the expected set of convergence and divergence patterns. References to the same flower within a game converged both to other descriptions used at that time point and to the post-game descriptions. This convergence was specific to flower and games, as references to different flowers within a game and references to the same flower across games both diverged.

@@

Our third consideration was how the different utility conditions moderated the reduction and convergence effects. We found some modest condition differences within games. In addition, groups in the shared utility condition diverged less from each other. This pattern of condition differences is consistent with less idiosyncratic descriptions in the shared utility condition.

@@

This dataset is just the start of testing how well findings from reference games generalize to more complex conversational situations. One dimension that we were unable to explore here due to the design of the game is the idea of shared versus privileged knowledge [@keysar2000taking; @heller2008role]. In the flowers game, like in reference games, all players saw the same set of images (and knew they did). In real-world situations, interlocutors may have different sets of objects in their context and may not know what is in their communication partner's context. An important future direction is studying how repeated reference phenomena interact with different communication goals and contexts.

@@

\setlength{\parindent}{-0.1in} 
\setlength{\leftskip}{0.125in}
\noindent
